---
title: "Assignment 2"
author: "Derek G. Nokes"
date: "Saturday, January 31, 2015"
output: pdf_document
toc: True
---

\pagebreak

# Connect to the PostgreSQL Database

To anwser assignment questions we connect to the PostgreSQL database as follows:

```{r}
#
library('RPostgreSQL')
# connect to PoststgreSQL using RPostgreSQL
driverHandle <- dbDriver("PostgreSQL")
 
# establish connection to 'flights' database
connectionHandle <- dbConnect(driverHandle, 
                              dbname="flights",
                              host="localhost",
                              port=5432,user="postgres",
                              password  ="tgdnrx78")
```

# Q1
Which destination in the flights database is the furthest distance away?

## Anwser

We join the 'flights' and 'airports' tables:

```{r}
# construct the query
# join flights and airports tables to match origin and destination
# FAA codes with airport names, latitude, and longitude
query<-'SELECT DISTINCT flights.origin as origin_faa,airports.name as 
origin_name,airports.lat as origin_lat,airports.lon as origin_lon,
flights.dest as dest_faa,a.name as dest_name,a.lat as dest_lat,a.lon 
as dest_lon,flights.distance FROM flights INNER JOIN airports ON 
flights.origin = airports.faa JOIN airports a ON flights.dest = a.faa 
GROUP BY origin_faa,dest_faa,flights.distance,origin_name,origin_lat,
origin_lon,dest_name,dest_lat,dest_lon ORDER BY flights.distance DESC 
LIMIT 1'

# fetch data
data<-dbGetQuery(connectionHandle,query)
knitr::kable(t(data), caption = 'Destination Furthest Distance Away')
```

`r data$dest_name` (`r data$dest_faa`) is the destination that is the 
furthest distance away from `r data$origin_name` (`r data$origin_faa`) 
at `r data$distance` miles.


# Q2 
What are the different numbers of engines in the planes table? For each 
number of engines, which aircraft have the most number of seats?

## Anwser

We can fetch the maxiumum number of seats for each distinct number of 
engines:

```{r}
# fetch the distinct number of engines and join on
query<-'SELECT DISTINCT planes.engines,MAX(planes.seats) as max_seats 
FROM planes GROUP BY engines ORDER BY engines DESC'
numberOfEngines<-dbGetQuery(connectionHandle,query)

knitr::kable(numberOfEngines, caption = 'Maximum Number of Seats by 
             Number of Engines')
```

We can then loop over each combination of number of engines and maximum number 
of seats and fetch the details of the corresponding planes:

```{r}
# 

for(i in seq_along(numberOfEngines[,1])) {
    numberOfEnginesAndSeats<-numberOfEngines[i,]
    engines<-numberOfEnginesAndSeats$engines
    seats<-numberOfEnginesAndSeats$max_seats
      
    # build the query
    query<-paste0('SELECT DISTINCT model,manufacturer,type,engines,seats 
                  FROM planes WHERE engines=',engines,' AND seats=',seats,
                  ' ORDER BY seats')
    # execute the query
    data<-dbGetQuery(connectionHandle,query)

    # create the results table
    print(data)
    knitr::kable(data)
}

```

We can also use a slightly more complex single query to return the plane 
details associated with the maximum number of seats for each number of 
engines group:

```{r}
# build the query
query<-'SELECT DISTINCT t.manufacturer,t.model,t.type,t.engine,t.engines,
t.seats FROM planes t INNER JOIN (SELECT DISTINCT engines as numberOfEngines,
max(seats) as max_seats FROM planes GROUP BY engines) group_t ON t.seats = 
group_t.max_seats AND t.engines = group_t.numberOfEngines ORDER BY t.engines 
DESC,t.seats DESC'

# execute the query
data<-dbGetQuery(connectionHandle,query)

# create the results table
knitr::kable(data, caption = 'Model with Maximum Number of Seats by Number 
             of Engines')
```

# Q3 
What weather conditions are associated with New York City departure delays?

## Anwser

We join the 'flights' and 'weather' tables so that we can examine the 
relationship between different weather variables and the departure 
delays:

```{r}
# build the query (join the weather and flights tables on date and hour)
query<-"SELECT w.w_date_hour,f.dep_delay,f.carrier,f.tailnum,f.flight,
f.dest,f.air_time,f.distance,f.origin,w.origin,w.temp,w.dewp,w.humid,
w.wind_dir,w.wind_speed,w.wind_gust,w.precip,w.pressure,w.visib FROM 
flights f INNER JOIN (SELECT concat_ws('-',weather.year,weather.month,
weather.day,weather.hour) as w_date_hour,weather.* FROM weather) w 
ON concat_ws('-',f.year,f.month,f.day,f.hour) = w.w_date_hour AND 
f.origin = w.origin ORDER BY f.dep_delay DESC"

# execute the query
data<-dbGetQuery(connectionHandle,query)
```

We extract each of the weather variables and create data frames for 
plotting:

```{r}
delayInHours<-data$dep_delay/60
delayIndex<-floor(delayInHours)>0 & floor(delayInHours)<5
tempDF<-data.frame(delayInHours[delayIndex],data$temp[delayIndex])
windSpeedIndex<-data$wind_speed<200 & floor(delayInHours)>0 & floor(delayInHours)<5
windSpeedDF<-data.frame(delayInHours[windSpeedIndex],data$wind_speed[windSpeedIndex])
windGustIndex<-data$wind_speed<200 & floor(delayInHours)>0 & floor(delayInHours)<5
windGustDF<-data.frame(delayInHours[windGustIndex],data$wind_gust[windGustIndex])
precipDF<-data.frame(delayInHours[delayIndex],data$precip[delayIndex])
pressureDF<-data.frame(delayInHours[delayIndex],data$pressure[delayIndex])
visibDF<-data.frame(delayInHours[delayIndex],data$visib[delayIndex])
```

We create 2D histograms to examine the relationship between departure 
delays and temperature, wind speed, wind gust, precipitation, pressure,
and visibility:

```{r,fig.width=6,fig.height=4.5,echo=FALSE}
library(hexbin)
# create the plot handle
h1 <- hexbin(tempDF,xlab='delay (in hours)',ylab='Temperature')
h2 <- hexbin(visibDF,xlab='delay (in hours)',ylab='Visibility')
h3 <- hexbin(windSpeedDF,xlab='delay (in hours)',ylab='Wind Speed')
h4 <- hexbin(windGustDF,xlab='delay (in hours)',ylab='Wind Gust')
h5 <- hexbin(precipDF,xlab='delay (in hours)',ylab='Precipitation')
h6 <- hexbin(pressureDF,xlab='delay (in hours)',ylab='Pressure')
```

There is no clear relationship between temperature and departure
delays, but high and low temperature appear as though they have 
more delays

```{r}
plot(h1,main='Temperature')
```

There is no clear relationship between visibility and departure
delays

```{r}
plot(h2,main='Visibility')
```

There is no clear relationship between wind speed and departure
delays

```{r}
plot(h3,main='Wind Speed')
```

There is no clear relationship between wind gust and departure
delays

```{r}
plot(h4,main='Wind Gust')
```

There is no clear relationship between precipitation and departure
delays

```{r}
plot(h5,main='Precipitation')
```

There is no clear relationship between pressure and departure
delays

```{r}
plot(h6,main='Pressure')
```

There are no clear relationships between departure delays and
any of the weather variables. This may be an artifact of the
way the table join was done. Rather than join on the departure
time, it might make more sense to join on the intended departure
time (i.e., the departure time adjusted for the delay)

# Q4 
Are older planes more likely to be delayed?

## Anwser

First we join the 'fights' and 'plane' tables and look at the delays
binned by hour. A delay in hours of zero indicates a delay of between 
1 minute and 59 minutes, a delay of one indicates a delay of between
1 hour and 1 hour and 59 minutes, and so forth:

```{r}
query<-"SELECT p.year,(f.dep_delay/60) as delay_in_hours FROM flights f 
JOIN (SELECT * FROM planes) p ON f.tailnum = p.tailnum WHERE f.dep_delay>0 
AND p.year >0 ORDER BY p.year"

# execute the query
data<-dbGetQuery(connectionHandle,query)

```

We create a 2D histogram to examine the relationship between departure 
delays and the age of the plane:

```{r}
h1<- hexbin(data,xlab='year',ylab='delay (in hours)')
plot(h1,main='Delay (In Hours) By Year of Aircraft')
```

There is no clear relationship between the age of the plane and the
aircraft model year.

```{r}
query<-"SELECT p.year,(f.dep_delay/60) as delay_in_hours FROM flights f 
JOIN (SELECT * FROM planes) p
ON f.tailnum = p.tailnum 
WHERE p.year >0
ORDER BY p.year"

# execute the query
flights<-dbGetQuery(connectionHandle,query)

```

We can see that there are considerably more flights for aircraft models 
of particular years:

```{r,fig.width=6,fig.height=6}
library('ggplot2')
qplot(factor(year), data=flights, geom="bar",
      xlab='total number of flights',ylab='year',
      main='Total Number of Flights By Aircraft Model 
      Year') + coord_flip()
```

There are far fewer total flights by very old planes.

When we group by aircraft model year and look at the delayed 
flights as a percentage of total flights, there is again no 
clear relationship

```{r}
query<-"SELECT p.year,COUNT(f.dep_delay) FROM flights f 
JOIN (SELECT * FROM planes) p ON f.tailnum = p.tailnum 
WHERE p.year >0 GROUP BY p.year
ORDER BY p.year"

# execute the query
flights<-dbGetQuery(connectionHandle,query)

query<-"SELECT p.year,COUNT(f.dep_delay) FROM flights f 
JOIN (SELECT * FROM planes) p ON f.tailnum = p.tailnum 
WHERE p.year >0 AND dep_delay > 0 GROUP BY p.year
ORDER BY p.year"

# execute the query
delays<-dbGetQuery(connectionHandle,query)

qplot(delays$year,delays$count/flights$count,xlab='year',
      ylab='% delayed',main='% of Flights Delayed By 
      Aircraft Model Year')

```

Theefore, the age of the aircraft does not consistently appear 
to cause more delays

# Q5
Ask (and if possible answer) a question that also requires joining information 
from two or more tables in the flights database, and/or assumes that additional 
information can be collected in advance of answering your question.

## Anwser

Are departure delays related to destination?

```{r}
# construct the query
# join flights and airports tables to match origin and destination
# FAA codes with airport names, latitude, and longitude
query<-'SELECT DISTINCT flights.dest as dest_faa,a.name as dest_name,
a.lat as dest_lat,a.lon as dest_lon,AVG(dep_delay) as delay,COUNT(dep_delay) 
as number_of_flights,AVG(flights.distance) as distance FROM flights INNER JOIN airports 
ON flights.origin = airports.faa JOIN airports a 
ON flights.dest = a.faa GROUP BY dest_faa,dest_name,dest_lat,dest_lon
ORDER BY number_of_flights,delay'

# fetch data
data<-dbGetQuery(connectionHandle,query)

```

Here is a plot of the destinations:

```{r}
library('rworldmap')
newmap <- getMap(resolution = "high")
plot(newmap,xlim = c(-139.3, -58.8),ylim = c(13.5, 55.7))
points(data$dest_lon, data$dest_lat, col = "red", cex = .6)
```

Here are the results

```{r}
knitr::kable(data, caption = 'Average Departure Delay By Destination')
```

There does not appear to be a relationship between the departure
delay and the destination distance.

```{r}
qplot(data$distance,data$delay,xlab='distance',
      ylab='delay',main='Delay By Destination Distance')
```

# Disconnect from the PostgreSQL Database

```{r}  
# Close PostgreSQL connection 
dbDisconnect(connectionHandle)

```


